Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-05-23 paper_jasa.tex
tweaking first two paragraphs of definitions/explanations

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{test_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots? *** Make the legend text use a subscript (MM: is it ok now?)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

\begin{table*}[hbtp]
\caption{Comparison of visual inference with existing inference technique. *** The plot for the traditional test is wrong, for two sided, p-value is area to the right and left of $|T(y|$. (MM: the vertical bar is showing the observed $T$ location not p-value.) }
\centering 
\begin{tabular}{llll} 
\hline
  & Traditional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

In general, let $\theta$ be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. Similarly, for any test, we have a rejection region $R$ and, complementary to that, $R^c$,  that partition the space of the test statistic, which is usually a subspace of the real numbers. For visual inference, the statistic is -- unlike in the classical hypothesis test -- not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the observed data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

\begin{dfn} \label{dfn:test}
A \textbf{visual test (statistic)} $T_\theta$ is defined as the grammar \citep{wilkinson:1999} of the plot,  consisting of type and specification of aesthetics necessary for complete reproducibility.
\end{dfn}

The visual test statistic is compared with values $ \theta \in \Theta_0$ using a lineup, which is defined as:

\begin{dfn}\label{dfn:lplot}
A \textbf{lineup}  is a layout of $m$ visual statistics, consisting of 
\begin{itemize}\itemsep-3pt
\item $m-1$ statistics simulated from the model specified by $H_0$  (null plots) and 
\item the test statistic produced by plotting the observed data, possibly arising from $H_1$.
\end{itemize}
\end{dfn}

The $(m-1)$ null plots can be considered to be samples drawn from the sampling distribution of the test statistic assuming that the null hypothesis is true.
If $H_1$ is true, \blue{we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots.}
{\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}  A careful visual inspection follows, and the observer is asked to point out the plot most different from the lineup.

Since the lineup plot consists of $m$ plots, the probability of choosing any one of them is $1/m$. Thus we have type-I error probability of $1/m$.

>>>>

\begin{figure}[htp]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{test_category.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the the largest shift or location difference between the boxplots?}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

\begin{table*}[hbtp]
\caption{Comparison of visual inference with existing inference technique.  }
\centering 
\begin{tabular}{llll} 
\hline
  & Traditional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that mapss the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T_0$ which would be possible values if $H_0$ was true.  %Similarly, for any test statistic, we have a rejection region $R$ and, complementary to that, $R^c$,  that partitions the space of the test statistic, which is usually a subspace of the real numbers. 
For visual inference, unlike in the classical hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the observed data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

\begin{dfn} \label{dfn:test}
A \textbf{visual test (statistic)} $T(y)$ is as a plot of the sample, and $T_0$ are plots of null data.
\end{dfn}

Ideally, the visual test statistic is defined using the grammar of graphics\citep{wilkinson:1999,hadley:2009}, consisting of type and specification of aesthetics necessary for complete reproducibility. The visual test statistic is compared with values $T_0$ using a lineup, which is defined as:

\begin{dfn}\label{dfn:lplot}
A \textbf{lineup}  is a layout of $m$ visual statistics, consisting of 
\begin{itemize}\itemsep-3pt
\item $m-1$ statistics, $T_0$ simulated from the model specified by $H_0$  (null plots) and 
\item the test statistic, $T(y)$ produced by plotting the observed data, possibly arising from $H_1$.
\end{itemize}
\end{dfn}

\noindent The $(m-1)$ null plots can be considered to be samples drawn from the sampling distribution of the test statistic assuming that the null hypothesis is true.
If $H_1$ is true, we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots. {\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}

A careful visual inspection of the lineup by an independent observer follows, and the observer is asked to point out the plot most different from the lineup.  Since the lineup plot consists of $m$ plots, the probability of choosing any one of them is $1/m$. Thus we have type-I error probability of $1/m$.
